Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-06-14 Paper/Paper.tex

@@

@@

I show summary statistics for the entire sample in Table \ref{tab:sumstats}. Only 3\% of physicians (approximately 770) retire over the course of the panel. 30\% of the physician-year observations have value one for working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting. By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure. Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015. The average physician is 45 years old, works with 1.5 hospitals and 1.3 systems.

@@

@@

I seek to estimate average treatment effects of EHR exposure on various physician outcomes. I treat EHR implementation as an exogenous treatment. In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues this causes in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}). Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}. I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}. These results yield similar patterns with slightly smaller magnitudes.

There are several assumptions necessary to identify $ATT(g,t)$. First, I assume that treatment is not reversed once it occurs. That is, once a physician is exposed to an EHR, they cannot be un-exposed. At the hospital level, this assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure. An EHR is a costly technology and requires a significant amount of collaboration to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a random sample of hospitals over time. Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR. Once they are exposed to the EHR for the first time, that experience cannot be forgotten. However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data. I investigate this possibility and limit the data to address it in Section \ref{sec:patientcount}.

@@

@@

@@

@@

@@

@@

\begin{figure}[htp]
    \centering
    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
    \caption*{Source: HealthIT.gov}
    \label{fig:meanuse}
\end{figure}

\begin{figure}[htp]
    \centering
    \caption{Screenshot of EHR System}
    \includegraphics[scale=.5]{Objects/epic-ehr-screenshot.jpg}
    \includegraphics[scale=.135]{Objects/EHRimage2.jpg}
    \label{fig:EPIC}
\end{figure}

\import{Table Code}{overall stats.tex}

\begin{figure}[p]
\centering
    \caption{Treatment Variables Over Time}
    \includegraphics[scale=.65]{Objects/sum_stats_year.pdf}
    \label{fig:treatmentgraph}
\end{figure}

\begin{figure}
    \centering
    \caption{Hospital EHR Adoption Over Time}
    \includegraphics[scale=.8]{Objects/hosp_treat.pdf}
    \label{fig:hosp_treat}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Retirement}
    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
    \label{fig:retirefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
    \label{fig:officefirst}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
    \label{fig:zip}
\end{figure}

\begin{figure}[p]
    \centering
    \caption{Effect of EHR Exposure on Patient Count}
    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
    \label{fig:patient}
\end{figure}

@@

Before I present these estimators, I present results from a two way fixed effects setup that may suffer bias from negative weighting. These results from each outcome are presented in

@@

\begin{figure}
    \centering
    \includegraphics{objects}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Limited Sample Years}
    \includegraphics[scale=.57]{Objects/years_graph.pdf}
    \label{fig:years}
\end{figure}

>>>>

\begin{figure}[ht]
    \centering
    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
    \caption*{Source: HealthIT.gov}
    \label{fig:meanuse}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Screenshot of EHR System}
    \includegraphics[scale=.4]{Objects/epic-ehr-screenshot.jpg}
    \includegraphics[scale=.115]{Objects/EHRimage2.jpg}
    \label{fig:EPIC}
\end{figure}

@@

I show summary statistics for the entire sample in Table \ref{tab:sumstats}. Only 3\% of physicians (approximately 770) retire over the course of the panel. 30\% of the physician-year observations have value one for working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting. By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure. Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015. The average physician is 45 years old, works with 1.5 hospitals and 1.3 systems. 
\import{Table Code}{overall stats.tex}

@@

\begin{figure}[t]
\centering
    \caption{Treatment Variables Over Time}
    \includegraphics[scale=.45]{Objects/sum_stats_year.pdf}
    \label{fig:treatmentgraph}
\end{figure}

@@

I treat EHR implementation as an exogenous treatment variable, and I seek to estimate average treatment effects of EHR exposure on various physician outcomes. In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues this causes in a classical two way fixed effects event study specification, I instead estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}). Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}. I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}. These results yield similar patterns with slightly smaller magnitudes.

Once I estimate the $ATT(g,t)$ for each group and year, I aggregate the estimates over groups to a more familiar event study plot. When the dis-aggregated estimates are of value, I present those as well. Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each table. Simultaneous confidence bands, calculated using bootstrapped standard errors, are presented.

\subsection{Assumptions}

There are several assumptions necessary to identify the parameters of interest, $ATT(g,t)$. First, I assume that treatment is not reversed once it occurs. That is, once a physician is exposed to an EHR, they cannot be un-exposed. At the hospital level, this assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure. An EHR is a costly technology and requires a significant amount of collaboration to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a 2\% random sample of hospitals over time\footnote{Note that out of the 81 hospitals included in the random sample, 4 of them show a reversal of treatment status. Upon further inspection, each of these hospitals show such a reversal due to a discrepancy in the coding of whether the hospital uses an EHR partially or fully. Those who use the EHR partially are coded as not fully implementing an EHR since it is not clear whether physicians in the sample would be exposed yet or not. Hospitals with this discretion arre coded as implementing an EHR in the first year that they answered "fully" to the survey question.}. Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR. Once they are exposed to the EHR for the first time, that experience cannot be forgotten. However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data. I investigate this possibility and limit the data to address it in Section \ref{sec:patientcount}.

\begin{figure}[ht]
    \centering
    \caption{Hospital EHR Adoption Over Time}
    \includegraphics[scale=.7]{Objects/hosp_treat.pdf}
    \label{fig:hosp_treat}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Effect of EHR Exposure on Retirement}
    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
    \label{fig:retirefirst}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0008), 0.002 (0.0009), 0.005 (0.002), respectively.}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
    \label{fig:officefirst}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0004 (0.008), -0.004 (0.008), 0.02 (0.02), respectively.}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
    \label{fig:officesecond}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0003 (0.004), -0.0007 (0.005), -0.006 (0.018), respectively.}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
    \label{fig:zip}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.045 (0.006), 0.046 (0.007), 0.042 (0.018), respectively.}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Effect of EHR Exposure on Patient Count}
    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
    \label{fig:patient}
    \vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 14.1 (4.6), 15.8 (5.4), -0.29 (13.0), respectively.}}
\end{figure}

@@

@@

Before I present these estimators, I present results from a two way fixed effects setup that may suffer bias from negative weighting. Specifically, I estimate the following equation:
$$\text{outcome}_{i,t}=\sum_{j=-4}^{-2} \text{rel\_year}_{j} + \sum_{j=0}^{4} \text{rel\_year}_{j} + \delta_i + \gamma_t,$$
where outcome$_{i,t}$ is one of the six outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to expansion, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
These results from each outcome are presented in Figure \ref{fig:twfe}. These results show similar patterns to the main specification, with slight variations in magnitude.

@@

\begin{figure}
    \centering
    \caption{Results: Two Way Fixed Effects}
    \includegraphics[scale=.8]{Objects/twfe_plot.pdf}
    \label{fig:twfe}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \caption{Limited Sample Years}
    \includegraphics[scale=.5]{Objects/years_graph.pdf}
    \label{fig:years}
\end{figure}
